One page description

### 2016-10-03 publications/Journal/3-features.tex
Small changes to section 3

Historically, \textit{dxex} is based on \textit{PythonPDEVS}~\cite{PythonPDEVS}.
Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
\textit{Dxex} is a C++11-based \textsf{Parallel DEVS} simulator, based on the design of \textit{PythonPDEVS}.
Whereas the feature set is not too comparable, the architectural design, simulation algorithms, and optimizations, are highly similar.

We will not make a detailed comparison with \textit{PythonPDEVS} here, but only mention some supported features.
\textit{Dxex} supports, similarly to \textit{PythonPDEVS}, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions~\cite{JDF}, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
We do not elaborate on these features in this paper.
But whereas \textit{PythonPDEVS} only supports optimistic synchronization, \textit{dxex} support multiple synchronization protocols (though only in parallel).
This is in line with the design principle of \textit{PythonPDEVS}: allow users to pass performance hints to the simulation kernel.
In our case, a user can pass the simulation kernel the synchronization protocol to use for this model, or even switch the synchronization protocol during runtime.
Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language. \textit{Dxex} will use new optimizations from the C++14 standard when possible.

@@

In the remainder of this section, we will elaborate on our prominent new feature: the efficient implementation of multiple synchronization protocols within the same simulation tool, which are offered transparently to the model.

\subsection{Synchronization protocols}
We previously explained the existence of different synchronization protocols, each optimized for a specific kind of model.
As no single synchronization protocol is ideal for all models, a general purpose simulation tool should support multiple protocols.
Currently, most parallel simulation tools choose only a single synchronization protocol due to the inherent differences between protocols.
We argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for \textit{dxex}.

These different protocols relate to three different model characteristics.
Conservative synchronization for when high lookahead exists between different nodes, and barely any blocking is necessary.
Optimistic synchronization for when lookahead is unpredictable, or there are rare (almost) instantaneous events.
Finally, sequential simulation is still required for models where parallelism is bad, where all protocols actually slow down simulation.

@@

\subsubsection{Conservative}
For conservative synchronization, each node must determine the nodes it is influenced by.
Each model needs to provide a lookahead function, which determines the lookahead depending on the current simulation state.
Within the returned time interval, the model promises not to raise an event.
A node aggregates this information to compute its Earliest Output Time (EOT).
This value is communicated to other nodes through shared memory.

Reading and writing to shared memory is done through the use of the new C++11 synchronization primitives.
Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
C++11 further allows us to make the implementation portable, as well as more efficient: the compiler makes further optimizations to heavily used components.

\subsubsection{Optimistic}
For optimistic synchronization, each node must be able to roll back to a previous point in time.
This is often implemented through the use of state saving.
This needs to be done carefully in order to avoid unnecessary copies, and minimize the overhead.
We use the default: explicitly save each and every intermediate state.
Mattern's algorithm~\cite{mattern} is used to determine the GVT, as it runs asynchronously and uses only $2n$ synchronization messages.
Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed, and irreversible actions are committed.

@@

\subsection{Synchronization Protocol Transparency}
We define synchronization protocol transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
User should thus only provide one model, implemented in C++11, which can be simulated using sequential execution, conservative synchronization, or optimistic synchronization.
The synchronization protocol to use is a simple configuration option.
The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.

@@

\subsubsection{Memory Management}
\label{sec:4-subsec:overhead-pgraph:memory}
\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
    \caption{Effect of memory allocators on parallel execution time.}
    \label{fig:memallocators_parallel}
\end{figure}
Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}: memory bandwidth doesn't increase as fast as the number of cores using it.
While this is always a problem, it is aggravated in \textit{dxex} by providing automatic memory management for events and states.
A model written for sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.

@@

Intra-kernel events can be (de)allocated without requiring synchronization with the other kernels. 
They can be returned immediately to the memory allocator as the lifetime of these objects is known at creation.
In contrast inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
As a result simulations with many inter-kernel events suffer a performance hit, whereas the impact of many intra-kernel events can be minimized using arena ~\cite{Arena} allocators.

\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
The latter can be faster, but at the cost of extra configuration.
The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}, which reduces lock contention in \texttt{malloc} calls.

@@

We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default Standard Template Library (STL) random number generator.
This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
\textit{Dxex}'s sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.

Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in sequential simulation is three times faster using TRNG compared to using the STL RNG.
For parallel simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between sequential and parallel simulation.
Conservative synchronization is almost insensitive to the changing of the RNG, though a slight increase in performance can be noted.
Optimistic synchronization slows down significantly when the RNG becomes part of the model state, since the state needs to be copied as well.
This becomes a significant overhead when using the STL RNG, since performance plummets to a fraction of the original.
Using TRNG avoids this problem completely, as the size of the RNG state is negligible.

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/rngspeedupeffectdevstone.eps}
    \caption{Speedup with different RNG usage patterns compared to STL random number generator.}
    \label{fig:Queuerngspeedup}
\end{figure}

>>>>

Historically, \textit{dxex} is based on \textit{PythonPDEVS}~\cite{PythonPDEVS}.
Python is a good language for prototypes, but performance has proven insufficient to compete with C++-based simulation kernels~\cite{MasterThesis}.
\textit{Dxex} is a C++11-based \textsf{Parallel DEVS} simulator, based on the design of \textit{PythonPDEVS}.
Whereas the feature set is not too comparable, the architectural design, simulation algorithms, and optimizations are highly similar.

We will not make a detailed comparison with \textit{PythonPDEVS} here, but only mention some supported features.
\textit{Dxex} supports, similarly to \textit{PythonPDEVS}, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions~\cite{JDF}, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
We do not elaborate on these features in this paper.
But whereas \textit{PythonPDEVS} only supports optimistic synchronization, \textit{dxex} support multiple synchronization protocols (though only in parallel).
This is in line with the design principle of \textit{PythonPDEVS}: allow users to pass performance hints~\cite{PythonPDEVS_ACTIMS} to the simulation kernel.
In our case, a user can configure the simulation kernel with the synchronization protocol to use, or even switch the synchronization protocol during runtime.
Our implementation in C++11 also allows for (compiled) optimizations which were plainly impossible in an interpreted language.
\textit{Dxex} uses new optimizations from the C++14 standard when possible.
The C++11 standard threading primitives are used to run the different simulation cores.
Within a single simulation core, OpenMP~\cite{openmp4} is used to parallelize the transition functions, as is usual in \textsf{Parallel DEVS}.

@@

In the remainder of this section, we elaborate on our prominent new feature: the efficient implementation of multiple synchronization protocols within the same simulation tool, offered transparently to the model.

\subsection{Synchronization protocols}
We previously explained the existence of different synchronization protocols, each optimized for a specific kind of model.
As no single synchronization protocol is ideal for all domains, a general purpose simulation tool should support multiple protocols.
We argue that a general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for \textit{dxex}.

These different protocols relate to three different model characteristics.
Conservative synchronization for when high lookahead exists between different nodes, and barely any blocking is necessary.
Optimistic synchronization for when lookahead is unpredictable, or there are rare (almost) instantaneous events.
Finally, sequential simulation is still required for models where parallelism is bad. 
It is possible for the synchronization overhead to become larger than the achieved parallelism, resulting in slower simulation than sequential.

Data exchange between different simulation cores happens through shared memory, using the new C++11 synchronization primitives.
Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
C++11 further allows us to make the implementation portable, as well as more efficient: the compiler makes further optimizations to heavily used components.

@@

\subsubsection{Conservative}
For conservative synchronization, each node must determine the nodes it is influenced by.
Each model needs to provide a lookahead function, which determines the lookahead depending on the current simulation state.
Within the returned time interval, the model promises not to raise an event.

\subsubsection{Optimistic}
For optimistic synchronization, each node must be able to roll back to a previous point in time.
This is often implemented through the use of state saving.
This needs to be done carefully in order to avoid unnecessary copies, and minimize the overhead.
We use the default: explicitly save each and every intermediate state.
Mattern's algorithm~\cite{mattern} is used to determine the GVT, as it runs asynchronously and uses only $2n$ synchronization messages.
Once the GVT is found, all nodes are informed of the new value, after which fossil collection is performed, and irreversible actions (\textit{e.g.}, printing) are committed.

@@

\subsection{Synchronization Protocol Transparency}
We define synchronization protocol transparency as having a single model, which can be executed on each supported synchronization kernel, without any modifications.
User should thus only provide one model, implemented in C++11, which can be simulated using sequential execution, conservative synchronization, or optimistic synchronization.
The synchronization protocol to use is a simple configuration option.
The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.

@@

\subsubsection{Memory Management}
\label{sec:4-subsec:overhead-pgraph:memory}
Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}, since memory bandwidth doesn't increase as fast as the number of cores using it.
While this is always a problem, it is aggravated in \textit{dxex} by providing automatic memory management for events and states.
A model written for sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.

@@

Intra-kernel events can be (de)allocated without requiring synchronization with the other kernels. 
They can be returned immediately to the memory allocator as the lifetime of these objects is known at creation.
In contrast, inter-kernel events need a GVT algorithm to determine when safe deallocation can occur, even in conservative synchronization.
As a result, simulations with many inter-kernel events suffer a performance hit, whereas the impact of many intra-kernel events can be minimized using arena allocators~\cite{Arena}.

\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
The latter can be faster, but at the cost of additional configuration.
The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}, reducing lock contention in \texttt{malloc} calls.

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
    %TODO: Add value on top of bars
    \caption{Effect of memory allocators on parallel execution time, using optimistic synchronization.}
    \label{fig:memallocators_parallel}
\end{figure}

@@

We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default Standard Template Library (STL) random number generator with optimistic synchronization.
This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
\textit{Dxex}'s sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.

Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in sequential simulation is three times faster using any of the three alternatives, only the STL RNG is slow.
For parallel simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between sequential and parallel simulation.
Conservative synchronization is almost unaffected by changing the RNG.

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/rngspeedupeffectdevstone.eps}
    %TODO: why is STL RNG in state just as fast in sequential? Does storing it in the state really do that much?
    %TODO: consistency: TRNG, STL RNG in state, and TRNG in state
    \caption{Speedup with different RNG usage patterns compared to STL random number generator.}
    \label{fig:Queuerngspeedup}
\end{figure}
